Composite Bayesian inference in hindsight

### 2024-03-11 papers/composite_inference/notes2024.tex
Some corrections in notes 2024

In the years 2015--2019, I worked on an attempt to put forward a non-conventional statistical inference paradigm, which I called composite Bayesian inference.

Composite Bayesian inference is a new viewpoint on the ``naive Bayes'' method that justifies it without relying on implausible independence assumptions. It also comes with a posterior distribution, which is a scaled version of the ``naive Bayes'' posterior, enabling the computation of conservative credibility sets.

Whether composite Bayesian inference is the right way to look at naive Bayes methods is questionable. But the interpretation it provides seems to be useful.

I tried to generalize it it using the concept of ``super composite likelihood'' function, where the weights assigned to the multiple atomic features are dependent on the variable to be inferred, {\em i.e.}, they are variable-of-interest dependent (VOID).

My ultimate goal here was to justify image registration similarity measures in a neat statistical inference framework.

I didn't do a great job at this. The super composite likelihood could more simply be derived as a regular composite likelihood using VOID features, without the need to introduce a more complex definition. The crux of the problem is to justify likelihood functions based on VOID features, but I didn't do that.

\subsection{VOID feature likelihood}

The idea of VOID feature-based likelihood was introduced by Baggenstoss in \cite{Baggenstoss-03}. It is definitely a relevant topic, but this initial work was not fully successful and didn't seem to convince many.

On the positive side, Baggenstoss gave a simple and general way, which he called ``PDF~projection theorem'', to construct a VOID feature-based likelihood function. As noted by \cite{Minka-04}, it is just an application of the maximum entropy, or minimum KL~divergence, principle.

However, it is easy to see how the PDF~projection method can lead to completely absurd likelihood functions, if the VOID features are chosen arbitrarily. Baggenstoss probably saw that but brushed the issue under the carpet: ``The choice of features remains an art requiring intuition''. This is nonsense. The choice of features should be driven by explicit assumptions.

@@

The least we can ask to the VOID feature likelihood is to be asymptotically consistent, that is, to be maximal at the true parameter value in the limit of large samples. If this property does not hold, then the method is worthwhile nothing.

A sufficient condition to guarantee consistency is that the VOID feature function $t_\theta$ statisfies:
$$
\forall (\theta, \theta'),
\quad
\int p(t_{\theta'}|\theta) \log \frac{p(t_{\theta'}|\theta)}{\pi(t_{\theta'})} dt_{\theta'}
  \leq
\int p(t_{\theta}|\theta) \log \frac{p(t_{\theta}|\theta)}{\pi(t_{\theta})} dt_{\theta}
.
$$

This condition may look complicated, but it only says that the VOID feature should always be the most informative about the variable of interest, among all feature candidates. In other words, $t_\theta$ should be, for any~$\theta$, the best statistic to test the hypothesis that~$\theta$ is the correct parameter vs the null hypothesis.

Let's take the same example as in my initial paper on composite inference. Assume we want to infer a disease category (infectious / cardiovascular) using VOID likelihood based on two simple clinical measures, body temperature and blood pressure. The above condition tells us that we need to associate body temperature with the infectious category, and blood pressure with the cardiovascular category. If we do it the other way round, the VOID feature likelihood will at best be useless, at worst biased.

@@

What I overlooked is that the likelihood extension I was considering, basically VOID feature likelihood, stems from a semi-generative statistical model. In other words, it's a partial model that doesn't fully describe images. Now, can consistency follow from partial modeling assumptions? It would be a nice property, but isn't it too much to ask?

@@

\subsection{Getting VOID feature likelihood right}

At the end of the day, the VOID feature likelihood method is just a restriction of Baggenstoss's 2003 proposal. We say: hey, this is a cool method but you can't choose VOID features randomly.

However, there are other aspects that differ.

First, we can apply MaxEnt in a broader context: instead of assuming the distributions $p(t_\theta|\theta)$ and $\pi(t_\theta)$, we can just assume some mean-value constraints and obtain the distributions via game theory. This is conceptually important as it can justify voxel independence assumptions.

Second, nuisance parameters are not correctly handled in Baggenstoss's work. He failed to see that the likelihood ratio involved in the PDF~projection fundamentally arises from a joint KL~minimization objective, hence it reflects a competition between two antagonist hypotheses. Parameter elimination should be consistent with that. This observation naturally leads to maximum likelihood ratio estimates rather than conventional likelihood parameter elimination.

>>>>

In the years 2015--2019, I worked on an attempt to put forward a slightly new statistical inference paradigm, which I called composite Bayesian inference.

Composite Bayesian inference is a new perspective on the ``naive Bayes'' method that justifies it without relying on implausible independence assumptions. It also comes with a posterior distribution, which is a scaled version of the ``naive Bayes'' posterior, enabling the computation of conservative credibility sets.

I do believe at this point that Bayesian inference is the right way to look at naive Bayes. At the very least, the interpretation it provides is insightful.

However, the most general version of composite inference I presented goes well beyond the classical na\"ive Bayes scope, and is based on the concept of ``super composite likelihood'' function, where the weights assigned to the multiple atomic features are dependent on the variable of interest, {\em i.e.}, they are based on variable data representations.

My hidden goal here was to justify image registration similarity measures in a neat statistical inference framework.

I didn't do a great job at this. The super composite likelihood could more simply be derived as a regular composite likelihood using variable features, without the need to introduce a more complex definition. The crux of the problem is to justify likelihood functions based on variable features, but I didn't do that.

\subsection{Variable feature likelihood}

The idea of variable feature (VF) based likelihood was introduced by Baggenstoss in \cite{Baggenstoss-03}. It is definitely a relevant idea, and perhaps an important one too, but this initial work was not fully successful.

On the positive side, Baggenstoss gave a simple and general way, which he called ``PDF~projection theorem'', to construct a VF-based likelihood function. As later noted by \cite{Minka-04}, it is just an application of the maximum entropy, or minimum KL~divergence, principle.

However, it is easy to see how the PDF~projection method can lead to heavily biased, hence completely absurd, likelihood functions, if the variable features are chosen arbitrarily. Fixed features can be chosen arbitrarily -- at worst, they will lead to a flat likelihood, but can't bias inference. The same cannot be said for variable features.

Baggenstoss probably saw that, but brushed the issue under the carpet: ``The choice of features remains an art requiring intuition''. This is nonsense. The choice of features should be driven by explicit assumptions.

@@

The least we can ask to the variable feature likelihood is to be weakly consistent, that is, to be maximal at the true parameter value in the limit of large samples. If this property does not hold, then the method is worthwhile nothing.

A sufficient condition to guarantee weak consistency is that the variable feature function $t_\theta$ statisfies:
$$
\forall (\theta, \theta'),
\quad
\int p(t_{\theta'}|\theta) \log \frac{p(t_{\theta'}|\theta)}{\pi(t_{\theta'})} dt_{\theta'}
  \leq
\int p(t_{\theta}|\theta) \log \frac{p(t_{\theta}|\theta)}{\pi(t_{\theta})} dt_{\theta}
.
$$

This condition may look complicated, but it only says that the variable feature should always be the most informative about the variable of interest, among all feature candidates. Specifically, $t_\theta$ should be, for any~$\theta$, the best statistic to test the hypothesis that~$\theta$ is the correct parameter vs the null hypothesis. In plain and simple words, the variable data representation should not discard any critical evidence.

Ideally, $t_\theta$ would be a sufficient statistic for each binary test $H_\theta$ vs $H_0$. Baggenstoss did note that, under sufficiency of $t_\theta$, the PDF~projection gives the true data distribution. But it is not a necessary condition for the method to work. And luckily so, otherwise the VF likelihood method would be pretty difficult to apply in practice.

Let's take the same example as in my initial paper on composite inference. Assume we want to infer a disease category using VF likelihood based on basic clinical measures, say, body temperature and blood pressure. The above condition tells us that we need to associate body temperature with infectious diseases, and blood pressure with the cardiovascular category. If we do it the other way round, the VF likelihood will at best be useless, at worst biased.

@@

What I overlooked is that the likelihood extension I was considering, basically VF likelihood, stems from a semi-generative statistical model. In other words, it's a partial model that doesn't fully describe images. Now, can consistency follow from partial modeling assumptions? It would be a nice property, but isn't it too much to ask?

@@

\subsection{Conservativeness of VF likelihood}

An interesting remark is that the expected ratio $\log p(t_\theta|\theta)/\pi(t_\theta)$ is lower than the true log-likelihood ratio $\log p(x|\theta)/\pi(x)$ by the general fact that the KL~divergence decreases under feature extraction. From there, we can argue that VF likelihood is conservative if used in a Bayesian framework.

\subsection{Getting VF likelihood right}

At the end of the day, the VF likelihood method is just a restriction of Baggenstoss's 2003 proposal. We say: hey, this is a cool method but you can't choose variable features randomly. Yet there are two other aspects in which my approach significantly deviated from Baggenstoss's.

\paragraph{Applying MaxEnt in a broader context.} Instead of assuming the distributions $p(t_\theta|\theta)$ and $\pi(t_\theta)$, we may just assume some mean-value constraints and derive epistemic distributions via KL~minimization, which brings connections with minimally informative likelihood, minimum description length, game theory and that kind of cool publishable stuff.

My motivation for this was primarily to justify voxel independence assumptions. However, this may be one of the worst ideas I had. In this more general framework, the key weak consistency property is gone -- think about it, do we want a theory that justifies cross-correlation for multimodal images? Moreover, we don't even need that to justify voxel independence for feature registration models: de Finetti's representation theorem does the job, as I noticed a while back (in 2005 I think). No need to generalize Baggenstoss's approach here: one point for him.

\paragraph{The treatment of nuisance parameters.} Baggenstoss failed to see that the likelihood ratio involved in the PDF~projection fundamentally arises from a joint KL~minimization objective, hence it reflects a competition between two hypotheses. Parameter elimination should be consistent with that. This observation naturally leads to maximum likelihood ratio estimates rather than conventional likelihood parameter elimination. Nuisance parameters are not correctly handled in Baggenstoss's work. That point is for me. I think.
